Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-02-14 response_letter.txt

* There are a couple of shortcuts in the notations that occasionally
distracted me (although it was correct). For instance, in definition 1,
point 3, e is an edge and X(v) is a set of vertice. Or in the definition
of E_R(S) the dependency between F and R is not explicited. Similarly,
definition 2 introduces the RNA design networks but the definition stays
at a relatively abstract level. If the object is associated with RNA, why
not using a specialized terminology (e.g. nucleotide instead of
variables)?

>>>>> Notation could be adapted; e.g. "nucleotide instead of
variables" would work for this paper (but not for all possible extensions)

* The experimental validation is limited, but it is (to my point-of-view)
not an issue as the core of the paper are the theoretical results. If
possible, it would be useful to expand this benchmark. For instance, by
including a discussion using known multi-stable RNAs and comparing real
and artificial sequences. It would be interesting to characterize which
parameters increase the difficulty of the design, or determine a priori
the range of possible target energies.

TODO Discuss:
Adenin-Riboswitch (Schwalbe). Details aus Stefans Diss: 3-state/4-state riboswitch.
Experimentell ueberpruefte Designs gibts bisher nur fuer 2-state.

@@

Controling the energy barrier between two stable structure An important
feature when designing multi-stable RNA

>>>>> INTERESTING IDEA; rather for design than sampling. We could try to optimize 
the direct path barrier in this way (for producing decoys...)

* It was not clear to me if the authors implemented dangling ends?

>>>>> NO; if possible, reformulate the resepctive sentence
CHECK

Typos:
* Eq 1: Z is a function of S but S is in the sum as well...
* Sec. 2, start of paragraph 2: "Given is"?
DONE

### 2018-02-15 RNARedPrint-Bioinfo2018.tex
formatting stuffs

\newcommand{\Def}[1]{{\it #1}}

@@

The \Def{dependency graph} of the set $\F$ of functions (on
sequences of length $n$) is the \Def{hypergraph}
$G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$. 
Our algorithms will critically rely on a \Def{tree decomposition} of the dependency graph, which we define below.
\begin{definition}[Tree decomposition and width]
  \label{def:treedecomp}
  Given a hypergraph $G=(X, E)$, a \Def{tree decomposition} is a
  pair $(T,\chi)$, where $T$ is an unrooted tree/forest, and
  $\chi(v)\subseteq X$ is the set of vertices from $G$ assigned to the node $v\in T$, such
  that
\begin{enumerate}
\item each $x\in X$ occurs in at least one $\chi(v)$;
\item for each $x\in X$, the set $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$.
\item for every hyper-edge $e\in E$ of $G$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
\end{enumerate}
The \Def{width} of a tree decomposition $(T,\chi)$ is defined as
$\width(T,\chi) = \min_{u\in T} |\chi(u)| - 1 $. The \Def{treewidth}
of $G$ is the smallest width in any tree decomposition of $G$.
\end{definition}

\section{Counting designs for multiple structures is \#{\sf P}-hard}\label{sec:counting}

@@

\begin{algorithm}
 \KwData{Node $u$, partial assignment $\val\in\assignments(\separator{u}{v})$;\newline
 Cluster tree $(T,\chi,\phi)$ and partition functions $\Message{u'}{v'}[\val']$, $\forall (u'\to{}v')\in T$ and $\val'\in\assignments(\separator{u'}{v'})$.}
 \KwResult{Boltzmann-distributed random assignment for the subtree rooted at $u$, under the assumption of a partial assignment $\val$.}
 \Fn{\Sample$(u,\val;T,\chi,\phi,m)$}{
   $r \gets \Random(\evalfor{\Message{u}{v}}{\val})$\;
   \For{$\val'\in\assignments(\difference{u}{v})$}{

@@

\subsection{Encoding sequence features, constraints, and energy
  models.}\label{sec:energy_models}

@@

A further \Def{rejection step} is applied to the generated structures to retain only sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}. The rationale for a rejection approach follows from: 
i)\emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing Cupal \emph{et al}~\cite{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}). 
%Indeed, as shown in Figure~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
For base pair-based energy models and special compatibility graph
(paths, cycles\ldots) this property rigorously follows from analytic
combinatorics, see Bender-Richmond-Williamson~\cite{Bender1983} and
Drmota~\cite{Drmota1997}. In such cases, the expected number of
rejections before reaching the targeted energies remains constant when
$\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when
$\varepsilon=0$. The \GCb-content of designs can also be controlled,
jointly and in a similar way, as done in
\Software{IncaRNAtion}~\cite{Reinharz2013}.

@@

\subsection{Implementation targeting Turner energies and GC-content}
We implemented the Boltzmann sampling approach (Algorithms
\ref{alg:pf} and \ref{alg:sampling}), performing sampling for given
target structures and weights $\pi_1,\dots,\pi_n$; moreover on top,
multi-dimensional Boltzmann sampling (see
Section~\ref{sec:multiBoltzmann}) to target specific energies and
\GCb-content.  Our tool \ourprog{} evaluates energies according to the base
pair energy model or the stacking energy model, using parameters which
were trained to approximate Turner energies
(Appendix~\ref{appsec:modelparameters}).
%
%\paragraph{Implementation of multi-dimensional Boltzmann sampling}
%As suggested in Section~\ref{sec:multiBoltzmann}, we heuristically optimize weights to target specific energies and GC content in an iterative procedure (based on the core sampling algorithm).
%Specifically, energy weights are initialized as $w_\ell=e^{\frac{1}{RT}}$; the gc weight as $w_{gc}=1$. For each iteration, $\psi\cdot n$ sequences are sampled with the current weights. Then, we determine the mean energies $E(S;R_\ell)$ and the mean GC content over the sampled sequences. For the next iteration, the weights are updated by the rules $w_\ell = w_\ell \gamma^{mean(E(S;R_\ell))-E_\ell}$ and $w_{gc} = w_{gc}\frac{GC_{target}}{mean(GC)}$. During this procedure, we collect   in a set $P$ all the sequences exhibiting energies within $[E_1-\delta,E_1+\delta]\times [E_2-\delta,E_2+\delta]\ldots$ and and GC-content within $[GC_{target}-\idelta',GC_{target}+\delta']$. 
%If $|P| > n$ or an maximum amount of iterations is reached, the sequences in $P$ are returned.
%

@@

To illustrate our above strategy, we sampled $n=1\,000$ sequences
targeting ${\sf GC}_{target}=0.5$ and different Turner energies for
the three structure targets of an example
instance. Figure~\ref{fig:energydist}A illustrates how the Turner
energy distributions of the shown target structures can be accurately
shifted to prescribed target energies; for comparison, we plot the
respective energy distribution of uniformly and Boltzmann sampled
sequences. Figure~\ref{fig:energydist}B summarizes the targeting accuracy for the single structure energies over a larger set of instances
from the \texttt{Modena} benchmark. We emphasize that only the simple stacking energies
are directly targeted (at $\pm$10\% tolerance, expect for a few hard
instances). Due to our linear adjustment, we still achieve well
controlled energies in the ultimately relevant Turner energy model.
\begin{figure*}[t]
  \begin{center}
    {\sf \bfseries A}\includegraphicstop[width=0.57\textwidth]{Figs/energy_distribution}
    {\sf \bfseries B}\includegraphicstop[width=0.32\textwidth]{Figs/offset}
  \end{center}
  \caption{%
  	Targeting specific energies using multi-dimensional Boltzmann sampling. (A) Turner energy distributions for three target structures (right) while targeting $(-40,-40,-20)$ and $(-20,-20,-20)$ free energy vectors. For comparison, we show the distribution of uniform and Boltzmann samples; respectively associated with homogenous weights $1$ and $e^\beta$. (B) Accuracy of targeting over all target structures of the \Software{Modena} benchmark instances \texttt{RNAtabupath} (2str), \texttt{3str} and \texttt{4str}. Shown is the relative deviation of targeted and achieved energies in the simple stacking model and the Turner model.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
}
  \label{fig:energydist}
\end{figure*}
\subsection{Generating high-quality seeds for further optimization}
We empirically study the capacity of \ourprog{} to improve the generation of seed sequences for subsequent local optimizations, showing an important application in biologically relevant scenario.

For a multi-target design instance, individual reasonable target energies are estimated from averaging the energy over samples at relatively high weights $e^{\beta}$, controlling the \GCb-content at 50\%. For each instance of the \Software{Modena} benchmark~\cite{Taneda2015}, we estimated such target energies and subsequently generated $1\,000$ seed sequences at these energies.
As its rationale, this procedure is tailored to produce sequences with similar Turner energy that favor the stability of the target structures having moderate \GCb-content; all these properties are desirable objectives for RNA design.
All sampled sequences are evaluated under the multi-stable design objective function of \cite{Hammer2017}:
\vspace{-12pt}
% \begin{align}
%   f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
%    & +\ \xi \frac{2}{k(k-1)} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
% \end{align}
\begin{align}
  \label{eq:blueprintobjective}
    f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
   & +\ 0.5\frac{1}{\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
\end{align}

@@

\newpage 
\section*{Acknowledgements}
YP is supported 
%jointly supported 
by
% the French 
{\em Agence
  Nationale de la Recherche} and the 
%Austrian
FWF (ANR-14-CE34-0011; project RNALands).  SH is supported by the
German Federal Ministry of Education and Research (BMBF support code
031A538B; de.NBI: German Network for Bioinformatics Infrastructure)
and the
% acknowledged by
%financial support 
of the Future and Emerging Technologies programme
% within the Seventh Framework Programme for Research of the
%European Commission, under 
(FET-Open grant 323987; project RiboNets).
%
We thank Leonid Chindelevitch for pointing out the potential for
complexity reductions due to the bipartiteness of the compatibility
graph; moreover, Arie Koster for discussing the practical computation
of the tree decomposition.

@@

In addition, we generated four sets of instances of increasing
complexity. The instances of the sets RF3, RF4, RF5, and RF6, each
respectively specify 3,4,5, and 6 target structures for sequence
length 100.  For each instance (100 instances per set), we generated a
set of $k$ ($k=3,\dots,6$) compatible structures as follows
\begin{itemize}
\item Generate a random sequence of length 100;
\item Compute its minimum free energy structure (ViennaRNA package);
\item Add the new structure to the instances if the resulting compatibility graph is bipartite;
\item Repeat until $k$ structures are collected.
\end{itemize}
For each instance, we generated the dependency graphs in the base pair
model and in the stacking model. Then, we performed tree decomposition
(using strategy ``GreedyFillIn'' of LibTW~\cite{Dijk2006}) on each dependency
graph. The obtained treewidths are reported in
Fig.~\ref{fig:td-widths}, while Fig.~\ref{fig:td-times} shows the
corresponding run-times of the tree decomposition.
Figure~\ref{fig:td-example} shows the tree decompositions for an example
instance from set RF3.

@@

For validating the trained parameters, we generate a second
independent test set of random RNA sequences in the same
way. Figure~\ref{fig:training-cor} shows correlation plots for the
trained parameters in the base pair and stacking models for predicting
the Turner energies in the test set with respective correlations of
$0.95$ an $0.94$.

>>>>

\newcommand{\Def}[1]{{\bfseries #1}}

@@

The \Def{dependency graph} of the set $\F$ of functions (on
sequences of length $n$) is the \Def{hypergraph}
$G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$. 
Our algorithms will critically rely on a \Def{tree decomposition} of the dependency graph, which we define below.
\begin{definition}[Tree decomposition and width]
  \label{def:treedecomp}
  Given a hypergraph $G=(X, E)$, a \Def{tree decomposition} is a
  pair $(T,\chi)$, where $T$ is an unrooted tree/forest, and
  $\chi(v)\subseteq X$ is the set of vertices from $G$ assigned to the node $v\in T$, such
  that
\begin{enumerate}
\item each $x\in X$ occurs in at least one $\chi(v)$;
\item  $\forall x\in X$, $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$;
\item $\forall e\in E$ of $G$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
\end{enumerate}
The \Def{width} of a tree decomposition $(T,\chi)$ is defined as
$\width(T,\chi) = \min_{u\in T} |\chi(u)| - 1 $. The \Def{treewidth}
of $G$ is the smallest width in any tree decomposition of $G$.
\end{definition}
\section{Counting valid designs is \#{\sf P}-hard}\label{sec:counting}

@@

\begin{algorithm}
 \KwData{Node $u$, partial assignment $\val\in\assignments(\separator{u}{v})$;\newline
 Cluster tree $(T,\chi,\phi)$ and partition functions $\Message{u'}{v'}[\val']$, $\forall (u'\to{}v')\in T$ and $\val'\in\assignments(\separator{u'}{v'})$.}
 \KwResult{Boltzmann-distributed random assignment for the subtree rooted at $u$, assuming a partial assignment $\val$.}
 \Fn{\Sample$(u,\val;T,\chi,\phi,m)$}{
   $r \gets \Random(\evalfor{\Message{u}{v}}{\val})$\;
   \For{$\val'\in\assignments(\difference{u}{v})$}{

@@

\subsection{Sequence features, constraints, and energy
  models.}\label{sec:energy_models}

@@

A further \Def{rejection step} is applied to the generated structures to retain only sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}. The rationale for a rejection approach follows from: 
i)\emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing Cupal \emph{et al}~\cite{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}). 
%Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
For base pair-based energy models and special compatibility graph
(paths, cycles\ldots) this property rigorously follows from analytic
combinatorics, see Bender-Richmond-Williamson~\cite{Bender1983} and
Drmota~\cite{Drmota1997}. In such cases, the expected number of
rejections before reaching the targeted energies remains constant when
$\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when
$\varepsilon=0$. The \GCb-content of designs can also be controlled,
jointly and in a similar way, as done in
\Software{IncaRNAtion}~\cite{Reinharz2013}.

@@

\subsection{Targeting Turner energies and GC-content}
We implemented the Boltzmann sampling approach (Algorithms
\ref{alg:pf} and \ref{alg:sampling}), performing sampling for given
target structures and weights $\pi_1,\dots,\pi_n$; moreover on top,
multi-dimensional Boltzmann sampling (see
Section~\ref{sec:multiBoltzmann}) to target specific energies and
\GCb-content.  Our tool \ourprog{} evaluates energies according to the base
pair energy model or the stacking energy model, using parameters which
were trained to approximate Turner energies
(Appendix~\ref{appsec:modelparameters}).
%
%\paragraph{Implementation of multi-dimensional Boltzmann sampling}
%As suggested in Section~\ref{sec:multiBoltzmann}, we heuristically optimize weights to target specific energies and GC content in an iterative procedure (based on the core sampling algorithm).
%Specifically, energy weights are initialized as $w_\ell=e^{\frac{1}{RT}}$; the gc weight as $w_{gc}=1$. For each iteration, $\psi\cdot n$ sequences are sampled with the current weights. Then, we determine the mean energies $E(S;R_\ell)$ and the mean GC content over the sampled sequences. For the next iteration, the weights are updated by the rules $w_\ell = w_\ell \gamma^{mean(E(S;R_\ell))-E_\ell}$ and $w_{gc} = w_{gc}\frac{GC_{target}}{mean(GC)}$. During this procedure, we collect   in a set $P$ all the sequences exhibiting energies within $[E_1-\delta,E_1+\delta]\times [E_2-\delta,E_2+\delta]\ldots$ and and GC-content within $[GC_{target}-\idelta',GC_{target}+\delta']$. 
%If $|P| > n$ or an maximum amount of iterations is reached, the sequences in $P$ are returned.
%

@@

To illustrate our above strategy, we sampled $n=1\,000$ sequences
targeting ${\sf GC}_{target}=0.5$ and different Turner energies for
the three structure targets of an example
instance. Fig.~\ref{fig:energydist}A illustrates how the Turner
energy distributions of the shown target structures can be accurately
shifted to prescribed target energies; for comparison, we plot the
respective energy distribution of uniformly and Boltzmann sampled
sequences. Fig.~\ref{fig:energydist}B summarizes the targeting accuracy for the single structure energies over a larger set of instances
from the \texttt{Modena} benchmark. We emphasize that only the simple stacking energies
are directly targeted (at $\pm$10\% tolerance, expect for a few hard
instances). Due to our linear adjustment, we still achieve well
controlled energies in the ultimately relevant Turner energy model.
\begin{figure*}[t]
  \begin{center}
    {\sf \bfseries A}\includegraphicstop[width=0.57\textwidth]{Figs/energy_distribution}
    {\sf \bfseries B}\includegraphicstop[width=0.32\textwidth]{Figs/offset}
  \end{center}
  \caption{%
  	Targeting specific energies using multi-dimensional Boltzmann sampling. (A) Turner energy distributions for three target structures (right) while targeting $(-40,-40,-20)$ and $(-20,-20,-20)$ free energy vectors. For comparison, we show the distribution of uniform and Boltzmann samples; respectively associated with homogenous weights $1$ and $e^\beta$. (B) Accuracy of targeting over all target structures of the \Software{Modena} benchmark instances \texttt{RNAtabupath} (2str), \texttt{3str} and \texttt{4str}. Shown is the relative deviation of targeted and achieved energies in the simple stacking model and the Turner model.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
}
  \label{fig:energydist}
\end{figure*}
\subsection{Generating high-quality seeds for further optimization}
We empirically study the capacity of \ourprog{} to improve the generation of seed sequences for subsequent local optimizations, showing an important application in biologically relevant scenario.

For a multi-target design instance, individual reasonable target energies are estimated from averaging the energy over samples at relatively high weights $e^{\beta}$, controlling the \GCb-content at 50\%. For each instance of the \Software{Modena} benchmark~\cite{Taneda2015}, we estimated such target energies and subsequently generated $1\,000$ seed sequences at these energies.
As its rationale, this procedure is tailored to produce sequences with similar Turner energy that favor the stability of the target structures having moderate \GCb-content; all these properties are desirable objectives for RNA design.
All sampled sequences are evaluated under the multi-stable design objective function of \cite{Hammer2017}:
% \begin{align}
%   f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
%    & +\ \xi \frac{2}{k(k-1)} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
% \end{align}
\begin{align}
  \label{eq:blueprintobjective}
    f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
   & +\ 0.5\frac{1}{\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
\end{align}

@@

\section*{Acknowledgements}
YP is supported 
%jointly supported 
by
% the French 
{\em Agence
  Nationale de la Recherche} and the 
%Austrian
FWF (ANR-14-CE34-0011; project RNALands).  SH is supported by the
German Federal Ministry of Education and Research (BMBF support code
031A538B; de.NBI: German Network for Bioinformatics Infrastructure)
and the
% acknowledged by
%financial support 
of the Future and Emerging Technologies programme
% within the Seventh Framework Programme for Research of the
%European Commission, under 
(FET-Open grant 323987; project RiboNets).
%
We thank Leonid Chindelevitch for pointing out the potential for
complexity reductions due to the bipartiteness of the compatibility
graph; moreover, Arie Koster for discussing the practical computation
of the tree decomposition.

@@

In addition, we generated four sets of instances of increasing
complexity. The instances of the sets RF3, RF4, RF5, and RF6, each
respectively specify 3,4,5, and 6 target structures for sequence
length 100.  For each instance (100 instances per set), we generated a
set of $k$ ($k=3,\dots,6$) compatible structures as follows
\begin{itemize}
\item Generate a random sequence of length 100;
\item Compute its minimum free energy structure (ViennaRNA package);
\item Add the new structure to the instances if the resulting compatibility graph is bipartite;
\item Repeat until $k$ structures are collected.
\end{itemize}
For each instance, we generated the dependency graphs in the base pair
model and in the stacking model. Then, we performed tree decomposition
(using strategy ``GreedyFillIn'' of LibTW~\cite{Dijk2006}) on each dependency
graph. The obtained treewidths are reported in
Fig.~\ref{fig:td-widths}, while Fig.~\ref{fig:td-times} shows the
corresponding run-times of the tree decomposition.
Fig.~\ref{fig:td-example} shows the tree decompositions for an example
instance from set RF3.

@@

For validating the trained parameters, we generate a second
independent test set of random RNA sequences in the same
way. Fig.~\ref{fig:training-cor} shows correlation plots for the
trained parameters in the base pair and stacking models for predicting
the Turner energies in the test set with respective correlations of
$0.95$ an $0.94$.
